Who Cares About Crackdowns? Exploring the Role of Trust in Individual Philanthropy

### 2020-10-03 manuscript/manuscript.Rmd
Final Suparna edits

---
title: "Who Cares About Crackdowns? Exploring the Role of Trust in Individual Philanthropy"
short-title: "Who Cares About Crackdowns?"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Marc Dotson
  affiliation: Brigham Young University
  url: https://marriottschool.byu.edu/directory/details?id=50683
  email: marc_dotson@byu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
date: "September 30, 2020"
published: Working paper.
git-repo: https://github.com/andrewheiss/who-cares-about-crackdown
abstract: >-
  Over the past two decades, international non-governmental organizations (INGOs) have faced increasing legal scrutiny and repression abroad. As a result, official aid to INGOs in repressive countries has reduced. NGOs thus need access to a diverse array of funds and may need to increase fundraising from private donors. However, research on the determinants of philanthropy towards international causes—particularly towards organizations working in challenging environments—is still in its infancy. Which demographic and experiential factors influence the decision of individuals to donate to legally besieged NGOs? We investigate this question using a market simulation based on a nationally representative conjoint experiment conducted on individuals across the U.S. with a history of charitable giving. We find that trust in political institutions and charities is enormously influential in shaping donor preferences. Donor profiles with high levels of social trust are likely to donate to human rights INGOs with friendly relationships with host governments, and that support holds steady if INGOs face criticism or crackdown. In contrast, donor profiles with lower levels of social trust prefer to donate to organizations that do not face criticism or crackdown abroad. Our findings have practical implications for INGOs raising funds from individuals amid closing civic space.
reference-section-title: References
toc: false
mainfont: Lora
sansfont: IBM Plex Sans
---

@@

In its 2016 annual report, Human Rights Watch claimed that civil society was under more aggressive attack than at any time in recent memory. Governments across the globe have cracked down on civil society organizations (CSOs) and non-governmental organizations (NGOs)—a phenomenon known as "closing civic space" [@CarothersBrechenmacher:2014; @CIVICUS:2017].[^NGOdef] As part of this phenomenon, NGOs face strong legal and bureaucratic obstacles to their work. Legal restrictions, or what we refer to as legal crackdowns, are a core part of these efforts. These crackdowns create barriers to entry, funding, and advocacy for NGOs in an effort to control, obstruct, and repress these organizations.

@@

Individual private donations are a critical source of INGO funding. Total private philanthropy in the United States has grown substantially over time, reaching more than \$400 billion in 2018 (@fig:giving-aggregate). In 2018, private giving to NGOs working in international affairs reached \$22.08 billion, and international affairs-focused organizations are one of only two categories of organizations that have continued experiencing substantial growth in giving [@givingusa:2019].[^intlaffairs] Most of this growth is driven by individuals making small-scale or micro-donations.[^smalldonors] In 2017, 70% of donations to international affairs NGOs came from individuals, rising to 86% in 2018 [@givingusa:2019]. But relying on individual donors means that NGOs must consider individual preferences when fundraising. Which individual donor characteristics and attributes are most salient for INGOs interested in maximizing philanthropy at a time when they are facing increasingly hostile environments abroad? What kinds of donors consider an INGO's host country political environment when making the decision to donate?

@@

We explore this question using a simulated market for philanthropic donations, based on data from a nationally representative sample of individuals in the United States who regularly donate to charity. We adapt methods used in marketing to generate dozens of distinct simulated personas that represent typical donor profiles, and we vary these personas along different dimensions of demographic, political, and social attributes. We find that specific individual characteristics help shape donor preferences for donating to INGOs, and that an organization's target issue area and its relationship with its host government interact strongly with these characteristics. Personas with high levels of trust in political and charitable institutions and who are actively engaged in charity and volunteerism prefer donating to human rights organizations that have friendly relationships with their host governments. In contrast, personas with low levels of social trust eschew both human rights organizations *and* any type of NGO with more contentious relationships with host governments, preferring instead to donate to NGOs with friendly host-country relationships and working on less contentious issues like emergency response and refugee relief. Personas with higher trust are more comfortable donating to legally besieged INGOs.

@@

We use a conjoint experiment to study the simultaneous effects of organizational characteristics and individual attributes on individuals' preferences to donate to a nonprofit organization. Conjoint analysis allows us to elicit donor preferences without respondents explicitly stating their exact preferences. This kind of analysis is commonly used in marketing to reveal consumer preferences for hypothetical products, and has become increasingly popular in political science and public policy [@KnudsenJohannesson:2019]. Because we are interested in exploring the complexity behind the decision to donate, we apply conjoint methods to the study of nonprofits and philanthropy.

@@

We begin by looking at differences in donation preferences across a range of basic demographic characteristics. The four panels of @fig:income show the average predicted market share for different organizational issue areas, funding sources, and relationships with host governments across different levels of persona income, with two different cutpoints for high and low income. To ease with modeling and interpretation, we split high and low income personas based on the 2017 U.S. median income (\$61,372), showing personas with incomes higher and lower than the national median. Because there is wide variation in donor incomes in the U.S., we also show the results from a simulation model with a more extreme split in income, holding annual persona income at \$50,000 and \$100,000+. The difference in the division of high and low income makes no difference in the market shares across issue areas and funding sources, but there are substantial income-based differences across different relationships with host governments.

@@

An organization's relationship with its host government does influence market shares across levels of income, but the size of this difference depends on the magnitude of income. When splitting income based on income, there is no difference between income levels. In general, personas prefer to donate to organizations with friendly relationships with their host governments and avoid those that are criticized or under crackdown. However, instead of looking at personas with below- and above-median incomes, we can simulate a subset of wealthier donors. These personas are more likely to prefer donating to organizations that are criticized by their host governments. This falls in line with their preference for human rights organizations, which have more contentious missions and are more likely to receive government criticism. This support, however, appears to evaporate as civil society crackdown intensifies. High income personas are the least likely demographic to prefer donating to NGOs under crackdown. Their lower income counterparts, on the other hand, are generally unfazed by NGOs' relationships with government, maintaining the same average market share across all three types of host country relationships. Patterns in preferences for issue area and funding sources remain constant across both methods of categorizing income—crackdown only has an effect on very high income personas. Income thus appears to influence preferences particularly for NGOs facing criticism and crackdown abroad—wealthy donors are perhaps fair-weather friends that are more likely to pull their support when an organization's work becomes more difficult.

@@

While education and religiosity are both important drivers of individual philanthropy, neither are associated with any sizable differences in persona donation patterns. As seen in @fig:edu-religion, preferences for issue areas are constant across both persona characteristics, with human rights and refugee relief organizations the most preferred, followed closely by emergency response organizations. regardless of levels of education or religiosity. Human rights issues attract a slightly larger market share of personas who are college graduates or who attend religious services at least monthly, while refugee issues are more likely to attract less religious personas, but these differences are minor. Preferences are consistent for different sources of funding as well—personas generally prefer organizations funded mostly by government grants. Persona preferences across different organizational relationships with host governments are also consisten across education and religiosity, with most personas preferring to donate to NGOs that enjoy friendly relations abroad. College graduate personas are slightly more willing to donate to organizations facing crackdown, but the difference is roughly half a percentage point.

@@

Given the nature of conjoint experiments, the randomly presented organizations may have influenced respondent choices. All four organizations are well-known, and it is likely that participant responses reflect prior beliefs and news coverage about Amnesty, Oxfam, Greenpeace, and the Red Cross. For instance, in February 2018, news broke out about sexual abuse and misconduct at Oxfam and several other prominent organizations, including Amnesty International and Save the Children. This potentially strengthened a narrative that the nonprofit sector suffers from governance failures [@Phillips:2019]. As such, it could be argued that respondents may be less likely to donate to Oxfam or Amnesty because of these scandals.

@@

Additionally, we remain skeptical that this scandal affected U.S. donors, given Oxfam is based in the UK, and "media narratives are often not as unsupportive as often assumed" [@BanksSchulpenBrockington:2020, 703]. Rather, lack of trust towards overseas NGOs, particularly aid and developmental agencies often results due to donors perceiving INGOs as having low personal relevance [@FaulknerTruongRomaniuk:2015]. We failed to find any systematic evidence that the Oxfam scandal influenced perceptions of the nonprofit sector in the U.S., likely because the story had far less news coverage in the U.S. Many individual donors have a preference to donate locally and they lack sufficient knowledge to make an informed decision about donating internationally. Overall, the literature suggests that for INGO fundraising, lack of awareness is a much larger issue than specific scandals or other knowledge of specific organizational characteristics that might reduce donor trust and confidence.

@@

The global crackdown on civil society and the consequent decline of official development assistance has made the role of philanthropy in sustaining INGOs clear. INGOs need access to multiple sources of funding, and the value of individual-level philanthropy is still relatively under-explored. While existing research has examined the impact of organizational-level factors on individual giving as well as giving to local causes, in this paper, we specifically look at the demographic traits of those who are likely to donate to INGOs facing hostile environments abroad. We find several important trends from our simulated market for philanthropy. On average, human rights organizations tend to attract personas who are wealthier, follow national and international news, and travel abroad. Donors who are less wealthy, follow the news less often, and do not travel are more attracted to organizations working on less contentious like emergency response and refugee relief.

More importantly, we find that trust in political institutions and in charitable organizations matters substantially in shaping donor preferences. Personas with high levels of social and charitable trust prefer donating to human rights organizations that have friendly relationships with their host governments, while personas with low trust are drawn more to emergency response and refugee relief organizations with unfriendly relationships with their host governments. Personas with low social trust tend to be wary of negative host-country relationships and strongly prefer to donate to organizations that are friendly with their host governments, while personas with high social trust are more willing to stick with legally besieged organizations. This could imply that donors that are the most trusting and involved could remain supportive when charities face government criticism and crackdown. The global crackdown on NGOs may also possibly sour NGOs' least trusting individual donors.

The above results have important implications for NGOs as they navigate a shrinking civic space and consider the use of different frames to tailor their fundraising appeals to individual donors. Our survey finds that at least for some kinds of individuals, NGOs may benefit from publicizing when they are targets of government crackdown, including that information in their fundraising appeals. However, the larger challenge may be addressing concerns of donors with low levels of trust who are the least likely to donate organizations facing criticism and crackdown abroad. More broadly, studies from a range of countries show that levels of giving are lowest for refugee and asylum organizations, as well as for overseas and development organizations. Rather, donors giving to international causes tend to support international disaster relief and religious groups [@RobsonHart:2020]. However, the former are precisely the organizations facing crackdown. More research is needed to answer how these organizations can appeal to broader array of potential foreign donors to increase levels of charitable-giving towards them.

>>>>

---
title: "Who Cares About Crackdowns? Exploring the Role of Trust in Individual Philanthropy"
short-title: "Who Cares About Crackdowns?"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Marc Dotson
  affiliation: Brigham Young University
  url: https://marriottschool.byu.edu/directory/details?id=50683
  email: marc_dotson@byu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
date: "October 2, 2020"
published: Working paper.
git-repo: https://github.com/andrewheiss/who-cares-about-crackdown
abstract: >-
  Over the past two decades, international non-governmental organizations (INGOs) have faced increasing legal scrutiny and repression abroad. As a result, official aid to INGOs in repressive countries has reduced. NGOs thus need access to a diverse array of funds and may need to increase fundraising from private donors. However, research on the determinants of philanthropy towards international causes—particularly towards organizations working in challenging environments—is still in its infancy. Which demographic and experiential factors influence the decisions of individuals to donate to legally besieged NGOs? We investigate this question using a market simulation based on a nationally representative conjoint experiment conducted on individuals across the U.S. with a history of charitable giving. We find that trust in political institutions and charities is enormously influential in shaping donor preferences. Donor profiles with high levels of social trust are likely to donate to human rights INGOs with friendly relationships with host governments, and that support holds steady if INGOs face criticism or crackdown. In contrast, donor profiles with lower levels of social trust prefer to donate to organizations that do not face criticism or crackdown abroad. Our findings have practical implications for INGOs raising funds from individuals amid closing civic space.
reference-section-title: References
toc: false
mainfont: Lora
sansfont: IBM Plex Sans
---

@@

In 2016, Human Rights Watch claimed that civil society was under more aggressive attack than at any time in recent memory [@Roth:2016]. Governments across the globe have cracked down on civil society organizations (CSOs) and non-governmental organizations (NGOs)—a phenomenon known as "closing civic space" [@CarothersBrechenmacher:2014; @CIVICUS:2017; @DPR:2015; @DPR:2016].[^NGOdef] As part of this phenomenon, NGOs face strong legal and bureaucratic obstacles to their work. Legal restrictions, or what we refer to as legal crackdowns, are a core part of these efforts. These crackdowns create barriers to entry, funding, and advocacy for NGOs in an effort to control, obstruct, and repress these organizations.

@@

Individual private donations are a critical source of INGO funding. Total private philanthropy in the United States has grown substantially over time, reaching more than \$400 billion in 2018 (@fig:giving-aggregate). In 2018, private giving to NGOs working in international affairs reached \$22.08 billion, and international affairs-focused organizations are one of only two categories of organizations that have continued experiencing substantial growth in giving [@givingusa:2019].[^intlaffairs] Most of this growth is driven by individuals making small-scale or micro-donations.[^smalldonors] In 2017, 70% of donations to international affairs NGOs came from individuals, rising to 86% in 2018 [@givingusa:2019]. But relying on individual donors means that NGOs must consider their preferences when fundraising. Which individual donor characteristics and attributes are most salient for INGOs interested in maximizing philanthropy at a time when they are facing increasingly hostile environments abroad? What kinds of donors consider an INGO's host country political environment when making the decision to donate?

@@

We explore this question using a simulated market for philanthropic donations, based on data from a nationally representative sample of individuals in the United States who regularly donate to charity. We adapt methods used in marketing to generate dozens of distinct simulated personas that represent typical donor profiles, and we vary these personas along different dimensions of demographic, political, and social attributes. We find that specific individual characteristics help shape donor preferences for donating to INGOs, and that an organization's target issue area and its relationship with its host government interact strongly with these characteristics. Personas with high levels of trust in political and charitable institutions and who are actively engaged in charity and volunteerism prefer donating to human rights organizations that have friendly relationships with their host governments. In contrast, personas with low levels of social trust eschew both human rights organizations *and* any type of NGO with more contentious relationships with host governments. Instead, they prefer donating to NGOs with friendly host-country relationships and working on less contentious issues like emergency response and refugee relief. Personas with higher trust are more comfortable donating to legally besieged INGOs.

@@

We use a conjoint experiment to study the simultaneous effects of organizational characteristics and individual attributes on individuals' preferences to donate to a nonprofit organization. Conjoint analysis allows us to elicit donor preferences without respondents explicitly stating their exact preferences. This kind of analysis is commonly used in marketing to reveal consumer preferences for hypothetical products, and has become increasingly popular in political science and public policy [@Helfer:206; @Mummolo:2016; @KnudsenJohannesson:2019]. Because we are interested in exploring the complexity behind the decision to donate, we apply conjoint methods to the study of nonprofits and philanthropy.

@@

We begin by looking at differences in donation preferences across a range of basic demographic characteristics. The four panels of @fig:income show the average predicted market share for different organizational issue areas, funding sources, and relationships with host governments across different levels of persona income, with two different cutpoints for high and low income. To ease with modeling and interpretation, we split high- and low-income personas based on the 2017 U.S. median income (\$61,372), showing personas with incomes higher and lower than the national median. Because there is wide variation in donor incomes in the U.S., we also show the results from a simulation model with a more extreme split in income, holding annual persona income at \$50,000 and \$100,000+. The difference in the division of high and low income makes no difference in the market shares across issue areas and funding sources, but there are substantial income-based differences across different relationships with host governments.

@@

An organization's relationship with its host government does influence market shares across levels of income, but the size of this difference depends on the magnitude of income. When splitting income based on income, there is no difference between income levels. In general, personas prefer to donate to organizations with friendly relationships with their host governments and avoid those that are criticized or under crackdown. However, instead of looking at personas with below- and above-median incomes, we can simulate a subset of wealthier donors. These personas are more likely to prefer donating to organizations that are criticized by their host governments. This falls in line with their preference for human rights organizations, which have more contentious missions and are more likely to receive government criticism. This support, however, appears to evaporate as civil society crackdown intensifies. High-income personas are the least likely demographic to prefer donating to NGOs under crackdown. Their lower income counterparts, on the other hand, are generally unfazed by NGOs' relationships with government, maintaining the same average market share across all three types of host country relationships. Patterns in preferences for issue area and funding sources remain constant across both methods of categorizing income—crackdown only has an effect on very high-income personas. Income thus appears to influence preferences particularly for NGOs facing criticism and crackdown abroad—wealthy donors are perhaps fair-weather friends that are more likely to pull their support when an organization's work becomes more difficult.

@@

While education and religiosity are both important drivers of individual philanthropy, neither are associated with any sizable differences in persona donation patterns. As seen in @fig:edu-religion, preferences for issue areas are constant across both persona characteristics, with human rights and refugee relief organizations the most preferred, followed closely by emergency response organizations, regardless of levels of education or religiosity. Human rights issues attract a slightly larger market share of personas who are college graduates or who attend religious services at least monthly, while refugee issues are more likely to attract less religious personas, but these differences are minor. Preferences are consistent for different sources of funding as well—personas generally prefer organizations funded mostly by government grants. Persona preferences across different organizational relationships with host governments are also consistent across education and religiosity, with most personas preferring to donate to NGOs that enjoy friendly relations abroad. College graduate personas are slightly more willing to donate to organizations facing crackdown, but the difference is roughly half a percentage point.

@@

The conjoint experiment conducted among a nationally representative sample of U.S. respondents with a prior history of charitable giving raises two main concerns for inference. It may be the case that respondent answers—and the corresponding beahavior of simulated personas—could potentially be driven by their priors towards these organizations. In addition, there are also important questions about the generalizability of these results outside the American context. We explore both below.

Given the nature of conjoint experiments, the randomly presented organizations may have influenced respondent choices. All four organizations are well-known, and it is likely that participant responses reflect prior beliefs and news coverage about Amnesty, Oxfam, Greenpeace, and the Red Cross. For instance, in February 2018, news broke out about sexual abuse and misconduct at Oxfam and several other prominent organizations, including Amnesty International and Save the Children. This may have potentially strengthened a narrative that the nonprofit sector suffers from governance failures [@Phillips:2019]. In a competitive NGO environment [@Cooley:2002], maintaining positive perceptions is important as donors who perceive these organizations as untrustworthy may simply take their resources elsewhere. As such, it could be argued that respondents may be less likely to donate to Oxfam or Amnesty because of these scandals.

@@

Additionally, we remain skeptical that this scandal affected U.S. donors, given Oxfam is based in the UK, and "media narratives are often not as unsupportive as often assumed" [@BanksSchulpenBrockington:2020, 703]. Moreover, research shows that when donors feel solidarity or an ideological connection with an NGO – especially because of the cause espoused by them – they are likely to consider the organization to be trustworthy, despite the lack of full transparency and oversight [@KeatingThrandardottir:2017]. Even watchdog organizations such as Charity Navigator and Better Business Bureau Giving Alliance have been found to not significantly impact individuals’ trust in nonprofit organizations. Rather, trust is more likely fostered through local networks and personal scrutiny [@TBPrakash:2017, 643].

We also failed to find any systematic evidence that the Oxfam scandal influenced perceptions of the nonprofit sector in the U.S., likely because the story had far less news coverage in the U.S. Many individual donors have a preference to donate locally and they lack sufficient knowledge to make an informed decision about donating internationally. @KnowlesSullivan:2017, conducting a field experiment in New Zealand, found that only a small number of participants in the study stated that not giving internationally was driven by mistrust. Rather, lack of trust towards overseas NGOs, particularly aid and developmental agencies often results due to donors perceiving INGOs as having low personal relevance [@FaulknerTruongRomaniuk:2015]. Overall, the literature suggests that for INGO fundraising, lack of awareness is a much larger issue than specific scandals or other knowledge of specific organizational characteristics that might reduce donor trust and confidence.

@@

The global crackdown on civil society and the consequent decline of official development assistance has made the role of philanthropy in sustaining INGOs clear. INGOs need access to multiple sources of funding, and the value of individual-level philanthropy is still relatively under-explored. While existing research has examined the impact of organizational-level factors on individual giving as well as giving to local causes, in this paper, we specifically look at the demographic traits of those who are likely to donate to INGOs facing hostile environments abroad. We find several important trends from our simulated market for philanthropy. On average, human rights organizations tend to attract personas who are wealthier, follow national and international news, and travel abroad. Donors who are less wealthy, follow the news less often, and do not travel, are more attracted to organizations working on less contentious like emergency response and refugee relief.

More importantly, we find that trust in political institutions and in charitable organizations matters substantially in shaping donor preferences. Personas with high levels of social and charitable trust prefer donating to human rights organizations that have friendly relationships with their host governments, while personas with low trust are drawn more to emergency response and refugee relief organizations with unfriendly relationships with their host governments. Personas with low social trust tend to be wary of negative host-country relationships and strongly prefer to donate to organizations that are friendly with their host governments, while personas with high social trust are more willing to stick with legally besieged organizations. This could imply who donors that are the most trusting and involved could remain supportive when charities face government criticism and crackdown. The global crackdown on NGOs may also possibly sour NGOs' least trusting individual donors.

The above results have important implications for NGOs as they navigate a shrinking civic space and consider the use of different frames to tailor their fundraising appeals to individual donors. Our simulation finds that at least for some kinds of personas, NGOs may benefit from publicizing when they are targets of government crackdown, including that information in their fundraising appeals. However, the larger challenge may be addressing the concerns of donors with low levels of trust, as they are the least likely to donate organizations facing criticism and crackdown abroad. More broadly, studies from a range of countries show that levels of giving are lowest for refugee and asylum organizations, as well as for overseas and development organizations. Rather, donors giving to international causes tend to support international disaster relief and religious groups [@RobsonHart:2020]. However, some of the former are precisely the organizations facing crackdown. More research is needed to answer how these organizations can appeal to broader array of potential foreign donors to increase levels of charitable-giving towards them.
